Mosaic_Benchmark_Figures

### 2025-04-04 analysis/manuscript_figures.Rmd
revised ortho analysis including statistical tests

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(patchwork)
library(viridis)
library(ggpubr)
library(ggupset)
```

@@

```{r}
(source_vaf_comp <- benchmark_source_comp_df %>%
    filter(
        str_detect(platform, "element|pacbio"),
        !str_detect(platform, "long"),
        !str_detect(platform, "sequel")
    ) %>%
    mutate(tech = str_remove(platform, "_.*")) %>%
    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
    summarise(VAF = sum(alt_count) / sum(depth)) %>%
    mutate(
        idx = paste0(CHROM, POS, REF, ALT),
        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
        tech = factor(tech, levels = c("element", "pacbio"), labels = c("Element", "PacBio Revio"))
    ) %>%
    ggpaired(
        x = "dna_source", fill = "tech",
        y = "VAF",
        id = "idx",
        facet.by = "tech",
        line.color = "gray",
        line.size = 0.4,
        palette = c("#7CAE00", "#C77CFF"),
    ) +
    stat_compare_means(paired = TRUE) +
    ylim(0, 0.4) +
    theme(legend.position = "none")
)
```

```{r}
ggsave(
    plot = source_vaf_comp,
    filename = here("figures", "fig4_VAF-RM-v-nonRM.pdf")
)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(patchwork)
library(viridis)
library(ggpubr)
library(ggupset)
library(glmmTMB)
library(broom)
```

@@

Preparing data and statistical tests

```{r}
## Preparing data
vaf_dat <- benchmark_source_comp_df %>%
    filter(
        str_detect(platform, "element|pacbio"),
        !str_detect(platform, "long|sequel")
    ) %>%
    mutate(
        platform = str_remove(platform, "_.*"),
        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
        variant_id = paste0(CHROM, ":", POS, ":", REF, ":", ALT)
    ) %>%
    group_by(dna_source, platform, variant_id) %>%
    summarise(
        vaf = sum(alt_count) / sum(depth),
        alt_count = sum(alt_count),
        ref_count = sum(ref_count),
        depth = sum(depth)
    )

# ---- Global GLMM Binomial Model ----
glmm_global <- glmmTMB(
    cbind(alt_count, ref_count) ~ dna_source + platform + (1 | variant_id),
    family = binomial,
    data = vaf_dat
)

# Overdispersion check
od_result <- performance::check_overdispersion(glmm_global)

# Reduced model (no dna_source)
glmm_null <- update(glmm_global, . ~ . - dna_source)
lrt_result <- anova(glmm_null, glmm_global)

# Extract LRT p-value
lrt_pval <- lrt_result$`Pr(>Chisq)`[2]

# ---- Per-Variant Binomial GLMs ----
variant_tests <- vaf_dat %>%
    group_by(variant_id) %>%
    filter(n_distinct(dna_source) == 2) %>%
    nest() %>%
    mutate(
        model = map(data, ~ glm(cbind(alt_count, ref_count) ~ dna_source + platform,
            family = binomial, data = .x
        )),
        tidy = map(model, ~ tidy(., conf.int = TRUE))
    ) %>%
    unnest(tidy) %>%
    filter(term == "dna_sourceRM") %>%
    mutate(p_adj = p.adjust(p.value, method = "BH"))

# Determine direction
variant_tests <- variant_tests %>%
    mutate(direction = case_when(
        p_adj < 0.05 & estimate > 0 ~ "Higher in RM",
        p_adj < 0.05 & estimate < 0 ~ "Higher in nonRM",
        TRUE ~ "Not significant"
    ))

# Count and get significant variants
total_sig <- sum(variant_tests$p_adj < 0.05)
higher_rm <- sum(variant_tests$direction == "Higher in RM")
higher_nonrm <- sum(variant_tests$direction == "Higher in nonRM")

sig_variants <- variant_tests %>%
    filter(p_adj < 0.05) %>%
    arrange(estimate) %>%
    mutate(row_id = row_number())
```

Out of `r nrow(variant_tests)` variants tested, `r total_sig` exhibited statistically significant VAF differences (FDR < 0.05) between RM and nonRM samples.
Of these, `r higher_rm` variants had higher VAFs in RM, while {higher_nonrm} had higher VAFs in nonRM.
The global GLMM model detected a significant overall difference in VAF profiles between RM and nonRM (LRT p = `r signif(lrt_pval, 3)`).
Overdispersion was not detected (dispersion ratio = `r signif(od_result$dispersion_ratio, 3)`, p = `r signif(od_result$p_value, 3)`).

#### Forest Plot - Supplemental?
```{r}
(forest_plot <- ggplot(
    sig_variants,
    aes(x = estimate, y = fct_reorder(variant_id, estimate))
) +
    geom_errorbarh(
        aes(
            xmin = conf.low,
            xmax = conf.high,
            color = direction
        ),
        height = 0.3
    ) +
    geom_point(color = "black") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(
        x = "Log-odds (RM vs nonRM)", y = "Variant ID",
        title = "Significantly Different VAFs (FDR < 0.05)"
    ) +
    scale_color_manual(values = c(
        "Higher in RM" = "red",
        "Higher in nonRM" = "blue"
    )) +
    theme_minimal()
)
```

Paried Plot
```{r}
# Merge direction info into your VAF comparison data
source_vaf_comp <- benchmark_source_comp_df %>%
    filter(
        str_detect(platform, "element|pacbio"),
        !str_detect(platform, "long|sequel")
    ) %>%
    mutate(tech = str_remove(platform, "_.*")) %>%
    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
    summarise(VAF = sum(alt_count) / sum(depth), .groups = "drop") %>%
    mutate(
        idx = paste0(CHROM, ":", POS, ":", REF, ":", ALT),
        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
        tech = factor(tech,
            levels = c("element", "pacbio"),
            labels = c("Element", "PacBio Revio")
        )
    ) %>%
    left_join(
        variant_tests %>%
            select(variant_id, direction),
        by = c("idx" = "variant_id")
    ) %>%
    mutate(
        line_color = case_when(
            direction == "Higher in RM" ~ "red",
            direction == "Higher in nonRM" ~ "blue",
            TRUE ~ "lightgrey"
        )
    )

# Generate ggpaired plot with custom line coloring
(ggpaired_plot <- ggpaired(
    source_vaf_comp,
    x = "dna_source", y = "VAF", id = "idx",
    facet.by = "tech",
    line.color = "direction", # use mapped line color
    line.size = 0.4
) +
    ylim(0, 0.4) +
    theme_minimal() +
    theme(legend.position = "none") +
    labs(
        title = "VAF Comparison by DNA Source and Platform",
        subtitle = "Line color indicates VAF direction for significant variants (FDR < 0.05)"
    )
)
```
